---
title: "P8131 HW5"
output: pdf_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS) # negative bino
library(pscl) # zero-inflated poission
library(tidyverse)
```

# 1. Study of nesting horseshoe crabs

## (a) Fit a Poisson model (M1) with log link with W as the single predictor. Check the goodness of fit and interpret your model.

```{r 1a_read, message=FALSE}
crabs <- read_table("HW5-crab.txt", col_names = TRUE)
crabs
```

### Fit a Poisson model (M1) with log link with W as the single predictor

```{r 1a_model, message=FALSE}
crabs.glm1 <- glm(Sa~W, family=poisson(link=log), data=crabs)
summary(crabs.glm1) # n=34, p=5-1 + 4-1 + 2-1 + 1=9
```

The model is $log(satellites_i) = \beta_1+\beta_2log(W_i) = -3.30476 + 0.16405 \cdot log(W_i)$

### Model Interpretation
for 

### Goodness of fit

Then we calculate chi-squared value to evaluate the goodness of fit of this model:

$H_0$: The model is close to the full model, $H_1$: not close to full model, significant level is 0.05

```{r 1a_goodness}
fit_p <- c(fitted.values(crabs.glm1))
# pearson residuals
pearsonresid <- (crabs$Sa-fit_p)/sqrt(fit_p)
# Chi-square X^2
chisq <- sum(pearsonresid*pearsonresid);chisq
deviance <- crabs.glm1$deviance;deviance
# test deviance
pchisq(deviance,df=crabs.glm1$df.residual, lower.tail = FALSE) # chisq test
```

So the Generalized Pearson Chisq Statistics is $\chi^2=544.157$, and deviance $D = 567.879$. The p-value of the goodness-of-fit chi-squared test is very small, meaning that the model does not fit the data well. We need to consider adding more predictors or dispersion parameter into the model. 

## (b) Fit a model (M2) with W and Wt as predictors. Compare it with the model in (a). Interpret your results.

### Fit M2
```{r 1b_model}
crabs.glm2 <- glm(Sa~W+Wt, family=poisson(link=log), data=crabs)
summary(crabs.glm2) # n=34, p=5-1 + 4-1 + 2-1 + 1=9
```

So the new model (M2) is $log(satellites_i) = \beta_1+\beta_2log(W_i)+\beta_3log(Wt_i) = -1.29168 + 0.04590 \cdot log(W_i) + 0.44744 \cdot log(Wt_i)$

### Model Interpretation

The model is interpreted as:

### Compare M1 and M2
```{r 1b_anova}
crabs.glm2$deviance
anova(crabs.glm1, crabs.glm2, test="Chisq")
```

The devaince of M2 is 559.8852, which is smaller than the deviance of M1. Also, the p-value of the chi-square test is $0.004695 < 0.05$, meaning that the predictor weight (Wt) is statistically significant in prediction of number of satellites. We can conclude that M2 better fits the data than M1. 

## (c) Check over dispersion in M2. Interpret the model after adjusting for over dispersion.

### Check for over dispersion in M2 via half-normal plot
```{r 1c_half}
plot(qnorm((173+1:173+0.5)/(2*173+1.125)),sort(abs(residuals(crabs.glm2,type='pearson',data=crabs))),
     xlab='Expected Half-Normal Order Stats',
     ylab='Ordered Abs Pearson Residuals')
abline(a=0,b=1)
```
The empirical slope is apparently off 1. Then we calculate the dispersion factor:

```{r 1c_overdisp}
crabs.glm3 <- glm(Sa~W+Wt, family=quasi(link=log,variance=mu), data=crabs)
summary(crabs.glm3) # check ?quasi
```

The Dispersion parameter is therefore 3.156449. 

### Model interpretation


# 2. Fish

## (a) Fit a Poisson model with log link to the data with area, year, and length as predictors. Interpret each model parameter.

```{r 2a_model}

```




