---
title: "Poisson Regression"
output: pdf_document
---

# Wave Damage Data (Page 205 of GLM)
The data concern a type of damage caused by waves to the forward section of certain cargo-carrying vessels. For the purpose of setting standards for hull construction, one needs to know the risk of damage associated with the three classifying factors shown below:

          Ship Type: A-E
Year of Construction: 1960-64, 65-69, 70-74, 75-79
Period of operation: 1960-74, 75-79

Two other variables are Aggregated months of service and Number of damage accidents.

The question of interest is how the number of damage accidents depends on the other four variables.

## formulate the data into a data.frame
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS) # negative bino
library(pscl) # zero-inflated poission
```

```{r}
month=c(127,63, 1095,1095,1512, 3353, 0,2244, 44882, 17176, 28609, 20370,  7064, 13099,0, 7117,1179,552,781,676,783,1948,0,274,251,105,
        288,192,349,1208, 0,2051,45,0,789,437,1157, 2161, 0,542)
damage=c(0 , 0  ,3  ,4  ,6, 18,  0, 11, 39, 29, 58, 53, 12, 44,  0, 18,  1,  1,  0,  1,  6,  2,  0,  1,  0,  0,  0,  0, 2, 11,  0,  4,  0,  0,  7,  7,  5, 12,  0,  1)
ship <- rep(LETTERS[1:5], rep(8,5)) # 5 groups
year <- rep(rep(c("60-64", "65-69", "70-74","75-79"), rep(2,4)), 5) # 4 groups
period <- rep(c("60-74", "75-79"), 20) # 2 groups
wave <- data.frame(ship, year, period, month, damage)
wave
```

## fit Poisson log linear model

```{r}
wave.glm1 <- glm(damage~ship+year+period+offset(log(month)), family=poisson, data=wave, subset=month>0)
summary(wave.glm1) # n=34, p=5-1 + 4-1 + 2-1 + 1=9
```

### check interactions 
```{r}
wave.glm2 <- glm(damage~ship*year+period+offset(log(month)), family=poisson(link=log), data=wave, subset=month>0)
summary(wave.glm2) # how to calc df?  
```

## deviance analysis (ignoring the over dispersion)
```{r}
test.stat=wave.glm1$deviance-wave.glm2$deviance
df=25-13
pval=1-pchisq(test.stat,df=df) # chisq test
pval # rej, go with the bigger model
```
rej, go with the bigger model


## estimate the dispersion parameter (from the additive model)
```{r}
# the traditional way of calc constant dispersion parameter
res.p1=residuals(wave.glm1,type='pearson',data=wave, subset=month>0)  # exactly the same as pearson residual for wave.glm3
G1=sum(res.p1^2) # calc dispersion param based on full model
pval=1-pchisq(G1,df=25);pval # lack of fit
phi=G1/(34-9);phi # 1.69
wave.glm1$deviance/wave.glm1$df.residual # dev/df, 1.55, greater than 1

summary(wave.glm1,dispersion=phi)
```

## an equivalent way of estimating dispersion parameter and fit over-dispersed poisson regression
```{r}
wave.glm3 <- glm(damage~ship+year+period+offset(log(month)), family=quasi(link=log,variance=mu), data=wave, subset=month>0)
summary(wave.glm3) # check ?quasi
```

## test over-dispersion (half normal plot)
```{r}
plot(qnorm((34+1:34+0.5)/(2*34+1.125)),sort(abs(res.p1)),
     xlab='Expected Half-Normal Order Stats',
     ylab='Ordered Abs Pearson Residuals')
abline(a=0,b=1)
abline(a=0,b=sqrt(phi),lty=2)  # controversial?
```

## F-test for deviance analysis of dispersion model
```{r}
test.stat=wave.glm1$deviance-wave.glm2$deviance # deviance (from original model fitting)
df=25-13
res.p=residuals(wave.glm2,type='pearson')  
res.p 
G=sum(res.p^2) # calc dispersion param based on larger model
phi=G/13
F.stat=test.stat/(df*phi)
pval=1-pf(F.stat,df,13)
pval # not rej, go with the smaller model
```

# negative binomial regression
```{r}
newwave=wave[wave$month>0,]
wave.nb=glm.nb(damage~ship+year+period+offset(log(month)),data=newwave)
summary(wave.nb) # dispersion param theta will be estimated simultaneously 
# the result (huge theta value) indicates no over-dispersion 
```


# Analyze fish data for Zero-inflated Poisson regression
```{r}
# install.packages("pscl")
zinb <- read.csv("https://stats.idre.ucla.edu/stat/data/fish.csv")
zinb=zinb[,c(3,4,5,8)]
head(zinb)
```

```{r}
m1 <- zeroinfl(count ~ child + camper | persons, data = zinb) # child and camper for poisson, persons for binary (event is true zero, aka not fishing)
summary(m1)
# interpretation:
# more people in the group, more likely to fish
# more children, fewer fish; those who camp tend to catch more fish
```

```{r}
pr <- predict(m1,type="zero")  # pi(do not fish)
mu <- predict(m1,type="count") # given fish, how many caught
zip <- pr + (1-pr)*exp(-mu) # total prob of catch 0 fish 
cbind(zip,zinb$count)
```

